Measurement Bias from Address Aliasing

### 2014-05-29 paper.tex
Fix references in introduction

The rest of this paper is structured as follows: Our methodology and experimental setup is explained in the next section, then section \ref{sec:aliasing} introduces ``4K aliasing'', providing necessary background for interpreting the later results.
Section \ref{sec:environment} presents an analysis of how the size of environment variables causes bias.
In Section \ref{sec:heap} we go through a similar analysis for bias to heap address alignment.
Related work is discussed in Section \ref{sec:related}.
Section \ref{sec:conclusions} includes a summary and conclusion.

\section{Experimental Methodology}
\label{sec:methodology}
In order to understand biased behavior it is important to be able to do precise and detailed measurements, without introducing \emph{observer effects}~\cite{Mytkowicz:2008:OE&MB}.
This can be achieved by using performance counters, instrumentation support in hardware that can be used to \emph{count} various events, such as cycles executed, branch misses, instructions fetched, and so on.
Recent Intel architectures have several hundred available events, providing a detailed view of what happens inside the CPU.
Performance counters are supported in the Linux kernel, accessed via a tool called \emph{perf}.
This tool instructs the kernel to enable the hardware Performance Monitoring Unit (PMU) for a different program, before it is executed.
Using this technique avoids modifications to the executed program, unlike tools like PAPI [LARS cite: PAPI: A Portable Interface to Hardware Performance Counters].
We use the perf-stat command in all experiments, which accepts raw PMU event codes listed in the reference manual~\cite{Volume3B}.

Reliability of PMU measurements have been evaluated by Weaver and Sally in [Can hardware performance counters be trusted? 2008].
The potential issues they find are either handled explicitly or are irrelevant to our setup.
%They find intra-machine variations of up to 1.07\%,

A small Python script is used to iterate over an exhaustive set of all known counters, which amounts to about 200 on our architecture.
Only a small set of events are collected at a time, to ensure events are actually counted continuously.
We do not use multiplexing between a limited set of counter registers, as it will introduce too many variations [Lars cite: Toward Accurate Performance Evaluation using Hardware Counters].
Controlled variations in environment size is performed by setting a dummy environment variable to $n$ number of zero characters, starting from a minimal environment.\footnote{Because perf-stat itself adds a few variables, the environment will never be completely empty.}
Interesting events are identified by computing linear correlation to cycle count, measuring all counters over a series of execution contexts.
% Cache related metrics are monitored in order to rule out cache as the underlying cause of bias, such as hit rates of micro-ops for each level of cache~\cite{OptimizationManual}.
Results are also averaged over multiple runs to reduce potential random error, using built-in perf rerun functionality.

@@

@@

\begin{narrowfig}{0.5\textwidth}[t]
  \centering
  \begin{tikzpicture}[font=\footnotesize]
    % See page 453 in pgf manual
    \node [
      rectangle split, rectangle split parts=10, 
      rectangle split part fill={white, white, lightgray, white, lightgray, white, lightgray, white},
      draw, anchor=center, text width=2cm
    ] (m)
      {
        \nodepart{one}
          environment
        \nodepart{two}
          stack
        \nodepart{four}
          mmap area
        \nodepart{six}
          heap
        \nodepart{eight}
          bss
        \nodepart{nine}
          data
        \nodepart{ten}
          text
      };
    \draw [decorate, decoration={brace, amplitude=5pt}] (m.south west) -- (m.seven split west) 
      node [black, midway, xshift=-1.5cm, text width=2.0cm] 
        { Program code and static data } ;
    \node[right] at (m.north east)
      { \scriptsize{0x7fff'ffffffff} } ;
    \node[right] at (m.south east)
      { \scriptsize{0x400000} } ;
  \end{tikzpicture}
  \caption{Relative positions of some important sections of memory at runtime, assuming a 64-bit process mapped to virtual memory.
  Initial addresses of stack, heap and memory mapped files are often randomized for security reasons.
  Addresses of code and statically allocated data are allocated at compile time by the linker, and can be determined by inspecting the executable.}
  \label{fig:virtualmemory}
\end{narrowfig}

>>>>

The rest of this paper is structured as follows: Our methodology and experimental setup is explained in the next section.
Then section \ref{sec:aliasing} introduces ``4K aliasing'', providing necessary background for interpreting the later results.
Section \ref{sec:environment} presents an analysis of how the size of environment variables causes bias, and in Section \ref{sec:heap} we go through a similar analysis for bias to heap address alignment.
Related work is discussed in Section \ref{sec:related}.
Section \ref{sec:conclusions} includes a summary and conclusion.

\section{Experimental Methodology}
\label{sec:methodology}
In order to understand biased behavior it is important to be able to do precise and detailed measurements, without introducing \emph{observer effects}~\cite{Mytkowicz:2008:OE&MB}.
This can be achieved by using performance counters, instrumentation support in hardware that can be used to \emph{count} various events, such as cycles executed, branch misses, instructions fetched, and so on.
Recent Intel architectures have several hundred available events, providing a detailed view of what happens inside the CPU.
Performance counters are supported in the Linux kernel, accessed via a tool called \emph{perf}.
This tool instructs the kernel to enable the hardware Performance Monitoring Unit (PMU) for a different program, before it is executed.
Using this technique avoids modifications to the executed program, unlike tools like PAPI~\cite{PAPI:PortableInterface}.
We use the perf-stat command in all experiments, which accepts raw PMU event codes listed in the reference manual~\cite{Volume3B}.

Reliability of PMU measurements have been evaluated by \citeN{Weaver:2008:PCTrusted}.
The potential issues they find are either handled explicitly or are irrelevant to our setup.
%They find intra-machine variations of up to 1.07\%,

A small Python script is used to iterate over an exhaustive set of all known counters, which amounts to about 200 on our architecture.
Only a small set of events are collected at a time, to ensure events are actually counted continuously.
We do not use multiplexing between a limited set of counter registers, as it will introduce too many variations~\cite{Wiplove:ImproveEstimation}. % Paper is rename of "Toward Accurate Performance Evaluation using Hardware Counters".
Controlled variations in environment size is performed by setting a dummy environment variable to $n$ number of zero characters, starting from a minimal environment.\footnote{Because perf-stat itself adds a few variables, the environment will never be completely empty.}
Interesting events are identified by computing linear correlation to cycle count, measuring all counters over a series of execution contexts.
% Cache related metrics are monitored in order to rule out cache as the underlying cause of bias, such as hit rates of micro-ops for each level of cache~\cite{OptimizationManual}.
Results are also averaged over multiple runs to reduce potential random error, using built-in perf rerun functionality.

@@

% Virtual memory/randomization illustration
\begin{narrowfig}{0.5\textwidth}[t]
  \centering
  \begin{tikzpicture}[font=\footnotesize]
    % See page 453 in pgf manual
    \node [
      rectangle split, rectangle split parts=10, 
      rectangle split part fill={white, white, lightgray, white, lightgray, white, lightgray, white},
      draw, anchor=center, text width=2cm
    ] (m)
      {
        \nodepart{one}
          environment
        \nodepart{two}
          stack
        \nodepart{four}
          mmap area
        \nodepart{six}
          heap
        \nodepart{eight}
          bss
        \nodepart{nine}
          data
        \nodepart{ten}
          text
      };
    \draw [decorate, decoration={brace, amplitude=5pt}] (m.south west) -- (m.seven split west) 
      node [black, midway, xshift=-1.5cm, text width=2.0cm] 
        { Program code and static data } ;
    \node[right] at (m.north east)
      { \scriptsize{0x7fff'ffffffff} } ;
    \node[right] at (m.south east)
      { \scriptsize{0x400000} } ;
  \end{tikzpicture}
  \caption{Relative positions of some important sections of memory at runtime, assuming a 64-bit process mapped to virtual memory.
  Initial addresses of stack, heap and memory mapped files are often randomized for security reasons.
  Addresses of code and statically allocated data are allocated at compile time by the linker, and can be determined by inspecting the executable.}
  \label{fig:virtualmemory}
\end{narrowfig}

@@
